References for PloS paper

### 2014-03-03 image_analysis_paper/math_description.tex
Updated math for paper

We will first discuss angular synchronization {\em only} considering rotations. 
%
Let $x_1, \dots, x_m \in \mathbb{R}^x$ denote the signals that we wish to align with respect to rotations.
%
We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
%
Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
%
Up to noise, 
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
 note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain {\em estimates} of $\theta_i - \theta_j$ by computing the rotations to optimally align $x_j$ to $x_i$, 
i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}

Rather than work with the angles $\theta_{ij}$ directly, it will be more convenient to work with the rotation matrices 
\begin{equation} \label{eq:R_theta}
R(\theta_{ij}) = \begin{bmatrix}
\cos(\theta_{ij}) & \sin(\theta_{ij}) \\
-\sin(\theta_{ij}) & \cos(\theta_{ij})
\end{bmatrix}
\end{equation}
%
Successive rotations correspond to multiplication of the rotation matrices, i.e., 
$R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$,
and, due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we describe this for general $d$ because we will later consider three--dimensional rotations).
%
We then construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij})
\end{equation}
%

We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, 
\begin{equation} 
H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
\end{equation}
 and so
\begin{equation} \label{eq:H_low_rank}
	H \approx 
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}
\end{equation}
%
We can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $\lambda_1 \ge \lambda_2 \ge \dots \ge \lambda_{md}$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$. 
%
Then
\begin{equation}
\hat{R} = 
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & | 
\end{bmatrix}
\end{equation}
where $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
%
Becaus $\hat{R}_i$ is not guaranteed to be a rotation matrix, we project $\hat{R}_i$ onto the closest rotation matrix. 
%
Let $U_i$ and $V_i$ denote the left and right singular vectors, respectively, of $\hat{R}_i$.
%
Then
\begin{equation}
R_{i, est} = U_i V_i^T
\end{equation} 
where $R_{i, est}$ is our estimate of $R(\theta_i)$. 
%
Note that $R_{i, est}$ is not guaranteed to be a proper rotation; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\rho_d$ so that (most of) the rotations are proper. 
%
We can estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating image $i$ by $-\theta_i$.

Furthermore, this eigendecomposition also considers {\em higher--order} consistency information. 
%
For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
\begin{equation} \label{eq:triplet_consistency}
R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
\end{equation}
should also hold.
%
However, 
\begin{equation}
(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj})
\end{equation}
and so {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
%
Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher--order considerations.

To illustrate how angular synchronization works, we have converted the two--dimensional images into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section). 
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile. 
%
The images are not aligned, and so rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization. 
%
The peaks are now (mostly) aligned.

@@

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
%
We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
%
In our framework, we assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
%
We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.

To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
%
We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
%
We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if $x_i$ is close to $x_j$,
\begin{equation}
W = 
\begin{bmatrix}
	1 & 1/2 & 0 & 0 & \dots & 0 \\
	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
	0 & 0 & 0 & \dots & 1 & 1/2 \\
	0 & 0 & 0 & \dots & 1/2 & 1 
\end{bmatrix}.
\end{equation}
%
If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \dots \cos((m-1) \pi / m) \cos(\pi)]^T$.
%
Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) < \phi_2(2) < \dots < \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
%
Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.

@@

We demonstrate how we can do this with our one--dimensional concentration profiles. 
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our distances in \eqref{eq:dmaps_W}.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) dmaps component, $\phi_2$.

@@

We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
%
Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image.
%
However, we now also obtain embedding coordinates for our images.
%
In general, the embedding coordinates are given by 
\begin{equation}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
%
If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e. the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
%
If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, in order to uncover the true manifold modulo symmetries.

@@

We are not only interested in registering the one--dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two--dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations. 
%
However, for \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
%
The typical matrix representation for the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
%
However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
%
We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Appendix).
%
Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
%
These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
%
See the Appendix for more details.

@@

The first step in angular synchronization or vector diffusion maps is to compute the translations and rotations to optimally align each pair of data points. 
%
For each image pair $I_i$ and $I_j$, we solve
\begin{equation}\label{eq:opt_pairwise}
(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
\begin{matrix}
\theta \in [0, 2\pi) \\ 
dx \in [-\Delta, \Delta]\\ 
dy \in [-\Delta, \Delta]
\end{matrix}
} \|g(I_j, \theta, dx, dy) - I_i \|_F^2.
\end{equation}
where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction. 
%
The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to interpolate the intensities of the pixels (since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image).
%
Also, the missing pixels in the rotated image due to corner effects are taken to be black.
%
The translations are done by shifting the pixels with periodic boundary conditions.
%
In \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image).
%
We only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.

The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution to \eqref{eq:opt_pairwise}.
%
We discritize the search space of rotations into 20 possible rotations ($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$), and 11 possible translations in both the $x$ and $y$ directions ($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$). 
%
We then check all possible combinations for the best rotation and translations that align $I_j$ to $I_i$. 
%
Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
%
The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
%
Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.

@@

\begin{figure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_1}
\includegraphics[width=\textwidth]{sphere2_1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_2}
\includegraphics[width=\textwidth]{sphere2_2}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_3}
\includegraphics[width=\textwidth]{sphere2_3}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_4}
\includegraphics[width=\textwidth]{sphere2_4}
\caption{}
\end{subfigure}
\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions. (a) The original image. (b) Rotation around the x--axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. (c) Rotation around the y--axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. (d) Rotation around the z--axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation. The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
\end{figure}

>>>>

We will first discuss angular synchronization {\em only} considering rotations. 
%
Let $x_1, \dots, x_m \in \mathbb{R}^n$ denote the signals that we wish to align with respect to rotations.
%
We assume that each signal $x_i$ is a noisy rotated copy of the underlying signal $x_{true}$ (which we are {\em not} given), such that 
\begin{equation}
x_i = f(x_{true}, \theta_i) + \xi_i
\end{equation}
where the function $f(x_{true}, \theta_i)$ rotates the signal $x_{true}$ by $\theta_i$ degrees, and $\xi_i$ is a noise term (often Gaussian noise). 
%
Our goal is to obtain estimates of $\theta_1, \dots, \theta_m$ so that we can align $x_1, \dots, x_m$. 
%
Up to noise, 
\begin{equation} \label{eq:pairwise_rot}
x_i \approx f(x_j, \theta_i - \theta_j) ;
\end{equation}
 note that \eqref{eq:pairwise_rot} does not require knowledge of $x_{true}$.
%
We can obtain an {\em estimate} of $\theta_i - \theta_j$ by computing the rotation that optimally aligns $x_j$ to $x_i$, 
i.e., $\theta_{ij} \approx \theta_i - \theta_j$, where
%
\begin{equation} \label{eq:opt_angle}
\theta_{ij} = \argmin_{\theta} \|x_i - f(x_j, \theta)\|^2.
\end{equation}

Rather than work with the angles $\theta_{ij}$ directly, it will be more convenient to work with the rotation matrices 
\begin{equation} \label{eq:R_theta}
R(\theta_{ij}) = \begin{bmatrix}
\cos(\theta_{ij}) & \sin(\theta_{ij}) \\
-\sin(\theta_{ij}) & \cos(\theta_{ij})
\end{bmatrix}
\end{equation}
%
Successive rotations now correspond to multiplication of rotation matrices, i.e., 
$R(\alpha_1 + \alpha_2) = R(\alpha_1) R(\alpha_2)$,
and, due to the orthogonality of rotation matrices, $R(-\alpha) = R(\alpha)^T$.

Let $d$ denote the dimension of the rotation matrices we are considering (for our example of planar rotations, $R(\theta_{ij}) \in \mathbb{R}^{2 \times 2}$ and $d=2$; we will write our procedure for general $d$ because we will later consider three--dimensional rotations).
%
We construct the matrix $H \in \mathbb{R}^{md \times md}$, where $H$ is an $m \times m$ matrix of $d \times d$ blocks with
\begin{equation} \label{eq:H_to_R}
H_{ij} = R(\theta_{ij})
\end{equation}
%

We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, 
\begin{equation} 
H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
\end{equation}
 and so
\begin{equation} \label{eq:H_low_rank}
	H \approx 
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
From \eqref{eq:H_low_rank}, we can see that we can (approximately) recover $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$ by computing the top block eigenvector of $H$, which we will denote $\hat{R}$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$. 
%
Then the top block eigenvector of $H$, $\hat{R}$, can be written as
\begin{equation}
\hat{R} = 
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & | 
\end{bmatrix},
\end{equation}
and $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
%
Because $\hat{R}_i$ is not guaranteed to be a rotation matrix, we must project $\hat{R}_i$ onto the closest rotation matrix \cite{...}. 
%
Let $U_i$ and $V_i$ denote the left and right singular vectors, respectively, of $\hat{R}_i$.
%
Then
\begin{equation}
R_{i, est} = U_i V_i^T
\end{equation} 
where $R_{i, est}$ is an orthonormal matrix and is our estimate of $R(\theta_i)$. 
%
Note that $R_{i, est}$ is not guaranteed to be a proper rotation (i.e. $det(R_{i, est} = \pm 1)$; because we are permitted to scale the eigenvectors of a matrix, we adjust the sign of $\\phi_d$ so that (most of) the rotations are proper. 
%
We can now estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and then register the signals by rotating signal $i$ by $-\theta_i$.

Furthermore, this formulation also considers {\em higher--order} consistency information. 
%
For example, given our pairwise estimates $R_{ij}$, we know that relationships of the form
\begin{equation} \label{eq:triplet_consistency}
R(\theta_{ik}) R(\theta_{kj}) \approx R(\theta_i) R(\theta_k)^T R(\theta_k) R(\theta_j)^T = R(\theta_i) R(\theta_j)^T
\end{equation}
should also hold.
%
Note that
\begin{equation}
(H^2)_{ij} = \sum_k R(\theta_{ik}) R(\theta_{kj});
\end{equation}
therefore, {\em all} infomation of the form in \eqref{eq:triplet_consistency} is contained in the matrix $H^2$.
%
Because $H$ and $H^2$ have the same eigenvectors, our problem formulation accounts for not only pairwise alignment information, but also these higher--order considerations.

To illustrate how angular synchronization works, we have converted the two--dimensional images shown in Figure \ref{fig:fluorescent_images} into concentration profiles on a ring (we will address the full two--dimensional images in a subsequent section). 
%
Figure \ref{subfig:1d_image_example} shows a typical image that we have converted to a one--dimensional concentration profile. 
%
The images are not aligned, and so rotation of this image corresponds to shifting (with periodic boundary conditions) the one--dimensional concentration profile. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows the concentration profiles for all the images in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one image.
%
Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles, now aligned using angular synchronization. 
%
The peaks are now (mostly) aligned.

@@

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we now wish to order them {\em in time} so that we can reconstruct the developmental dynamics.
%
We will use diffusion maps (DMAPS) \cite{coifman2005geometric} to temporally order our data.
%
DMAPS aims to uncover a parameterization of high-dimensional data sampled from a low-dimensional nonlinear manifold.
%
In our analysis, we will assume that our data lie on a one--dimensional curve or manifold, and that this one dimension is correlated with time. 
%
We will uncover a parameterization of this curve by only considering {\em local} neighbors of each data point.

To illustrate how local neighbors can allow us to uncover a global parameterization, assume we have data $x_1, x_2, \dots, x_m \in \mathbb{R}$ such that $x_1 < x_2 < \dots < x_m$.
%
We consider $x_i$ ``close to'' $x_{i+1}$ and $x_{i-1}$, and $x_i$ ``far away from'' all other data points. 
%
We then construct the (symmetric) weight matrix, $W \in \mathbb{R}^{m \times m}$,
\begin{equation}
W = 
\begin{bmatrix}
	1 & 1/2 & 0 & 0 & \dots & 0 \\
	1/2 & 1 & 1/2 & 0 & \dots & 0 \\
	\ddots & \ddots & \ddots & \ddots & \ddots & \ddots \\
	0 & 0 & 0 & \dots & 1 & 1/2 \\
	0 & 0 & 0 & \dots & 1/2 & 1 
\end{bmatrix}.
\end{equation}
%
$W_{ij}$ is large if $x_i$ is close to $x_j$,
%
If we normalize this matrix such that $\sum_j W_{ij} = 1$, then one can show that the first eigenvector of this normalized matrix is the all--ones vector, $\phi_1 = [1 1 \dots 1]^T$, and the second eigenvector is a cosine mode, $\phi_2 = [cos(\pi/m) \cos(2 \pi/ m) \cdots \cos((m-1) \pi / m) \cos(\pi)]^T$.
%
Therefore, the second eigenvector, $\phi_2$ is {\em one--to--one} with the parameterization of the data, i.e. $\phi_2(1) > \phi_2(2) > \dots > \phi_2(m)$, just as $x_1 < x_2 < \dots < x_m$. 
%
Therefore, sorting our data by the value of the entries in $\phi_2$ will allow us to order our data.

@@

We demonstrate how we can apply these techniques to our (aligned)one--dimensional concentration profiles. 
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our distances in \eqref{eq:dmaps_W}.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non--trivial) dmaps component, $\phi_2$.

@@

We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
%
Again, the top (block) eigenvector of $S$ contains approximations of the optimal rotations for each image, and the rotations can be computed as described in Section \ref{subsec:angsynch}.
%
However, the eigenvectors now also contain information about the embedding coordinates for our images.
%
In general, the embedding coordinates are given by 
\begin{equation}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
%
If we assume that our data are one--dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, one can show that a parameterization of our data (i.e. the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
%
If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, in order to uncover the true manifold modulo symmetries.

@@

We are not only interested in registering the one--dimensional concentration profiles with respect to rotational symmetries, but we would also like to register the two--dimensional images shown in Figure \ref{fig:fluorescent_images} with respect to rotations {\em and} translations. 
%
However, for \eqref{eq:H_low_rank} to be satisfied, we require that the matrices used to represent the symmetry group be orthogonal, i.e. $R(\alpha)^{-1} = R(\alpha)^T$. 
%
The typical matrix representation for the group of two--dimensional translations and rotations (using the matrix representation for the affine group \cite{...}) does not satisfy this property. 
%
However, we can (approximately) represent rotations and translations in two--dimensions using three--dimensional rotation matrices.
%
We do this by projecting the image onto a portion of the surface of a (three--dimensional) sphere (see Figure \ref{fig:SO3_picture}).
%
Rotations in two--dimensions now correspond to rotations around one principal axis in three--dimensions, and translations in two--dimensions correspond (approximately) to rotations around the other two principal axes in three--dimensions \footnote{Clearly, not all translations can be described this way, as translations can range from $-\infty$ to $+ \infty$, and rotations are only from $0$ to $2 \pi$. However, the images we will be considering are already mostly centered, and so we are only interested in small translations that are well within the $[0, 2\pi)$ range.}.
%
These rotation matrices are (by definition) orthoganol, and successive applications of various translations and rotations can be described via multiplication of the corresponding rotation matrices in $SO(3)$.
%
See the Appendix for more details.

@@

The first step in angular synchronization or vector diffusion maps is to compute the translations and rotations to optimally align each pair of data points. 
%
For each image pair $I_i$ and $I_j$, we solve
\begin{equation}\label{eq:opt_pairwise}
(\theta_{ij}, dx_{ij}, dy_{ij}) = \argmin_{
\begin{matrix}
\theta \in [0, 2\pi) \\ 
dx \in [-\Delta, \Delta]\\ 
dy \in [-\Delta, \Delta]
\end{matrix}
} \|g(I_j, \theta, dx, dy) - I_i \|_F^2.
\end{equation}
where $\| \cdot \|_F$ denotes the Frobenius norm, $\Delta$ is the maximum number of pixels by which we are allowed to translate the image, and $g(I_j, \theta, dx, dy)$ is a function that first rotates the image $I_j$ by $\theta$ degrees, then translates the image by $dx$ pixels in the $x$ direction, and finally translates the image by $dy$ pixels in the $y$ direction. 
%
The image rotation is done with the \texttt{imrotate} function in Matlab, using nearest neighbor interpolation to estimate the pixel intensities, since there is no direct mapping of the pixels of the unrotated image to the pixels of the rotated image.
%
The missing pixels in the rotated image due to corner effects are taken to be black.
%
The translations are done by shifting the pixels with periodic boundary conditions.
%
In \eqref{eq:opt_pairwise}, $\Delta$ is chosen such that only edge pixels with little to no signal are shifted, and the main image is not split or separated when translating (we take $\Delta=20$, which corresponds to a 20\% shift in the image).
%
We only consider shifts that correspond to an integer number of pixels, to remove the need for interpolation when doing the translations.

The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure to compute the solution to \eqref{eq:opt_pairwise}, we choose to discretize the search space and exhaustively search to estimate the solution.
%
We discritize the search space of rotations into 20 possible rotations ($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$), and 11 possible translations in both the $x$ and $y$ directions ($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$). 
%
We then check all possible combinations for the best rotation and translations that align $I_j$ to $I_i$. 
%
Although this can be somewhat time intensive, it is not prohibitive for the size of data sets which we are considering, and can be trivially parallelized for larger data sets if necessary.
%
The solution to this search will not be the exact solution to \eqref{eq:opt_pairwise}, but it will (most likely) be a close approximation.
%
Since our techniques are robust to noise, close approximations will be sufficient to obtain accurate results.

@@

\begin{figure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_1}
\includegraphics[width=\textwidth]{sphere2_1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_2}
\includegraphics[width=\textwidth]{sphere2_2}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_3}
\includegraphics[width=\textwidth]{sphere2_3}
\caption{}
\end{subfigure}
\begin{subfigure}{0.2\textwidth}
\includegraphics[width=\textwidth]{sphere_4}
\includegraphics[width=\textwidth]{sphere2_4}
\caption{}
\end{subfigure}
\caption{Illustration of how rotations in three dimensions correspond to translations and rotations in two dimensions. (a) The original image. (b) Rotation around the x--axis (Euler angle $\alpha$) in three dimensions corresponds to rotation of the image. (c) Rotation around the y--axis (Euler angle $\beta$) in three dimensions corresponds to horizontal translation. (d) Rotation around the z--axis (Euler angle $\gamma$) in three dimensions corresponds to vertical translation. The top row shows the three--dimensional spheres, and the bottom row shows the (two--dimensional) surface of the sphere in which we are interested.}
\label{fig:SO3_picture}
\end{figure}

As discussed in Section \ref{subsec:trans_rot_register}, the underlying symmetry group needs to have a real and orthoganol matrix representation. 
%
We therefore convert $ISO(2)$, the group of two--dimensional translations and rotations, to $SO(3)$, the group of three--dimensional rotations. 
%
In the space of three--dimensional rotations, the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$ correspond to
\begin{equation} \label{eq:angle_relations}
\begin{aligned}
	\alpha_{ij} &= \theta_{ij} \\
	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
\end{aligned}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere in $\mathbb{R}^3$.
%
From here, we can write rotations around the three principal axes in terms of the three Euler angles
\begin{equation}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix} 
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1 
    \end{bmatrix}
\end{aligned}
\end{equation}
where $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$ correspond to rotations around the $x$, $y$, and $z$ axes, respectively.
%
We then write the total rotation $R_{ij} \in SO(3)$ as 
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij})
\end{equation}
%
\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).
